Generation of TypeScript Declaration Files from JavaScript Code

### 2021-05-26 mplr-2021/main.tex
explaining globtoregexp

\section{Motivating Example}
\label{sec:motivating-example}
The \NPM{} package \texttt{glob-to-regexp} provides functionality to turn a
glob expression for matching filenames in the shell into a regular
expression\footnote{\url{https://www.npmjs.com/package/glob-to-regexp}}. It 
has about 9.9M weekly downloads and 188 \NPM{} packages depend on it. If
a developer creates or extends TypeScript code that depends on the
\texttt{glob-to-regexp} library, the TypeScript
compiler and IDE require a declaration file for that library to
perform static checking and code completion, respectively. With
\texttt{dts-generate} we automatically generate a TypeScript
declaration file for \texttt{glob-to-regexp}. Our tool downloads the \NPM{} 
package, runs the examples extracted from its documentation, gathers
run-time information, and generates a TypeScript declaration
file. The result is 
shown in \figref{fig:motivating-example-glob-to-regexp-vscode} and it is
ready for use in a TypeScript project:  Visual Studio's
code completion runs properly with the generated file. If the
\texttt{glob-to-regexp} package gets modified in the future, a new declaration
file can be generated automatically using
\texttt{dts-generate}. Our comparator tool (Section~\ref{sec:dts-compare}) can compare the new file
for incompatibilities with the previous declaration file.

@@

To construct the type signature for a function, we inspect the
primitive types of each argument as well as the interface created from
interactions on the argument in the function body. We do the same for
the return type of the function. This procedure yields a candidate
type for each function call observed during the execution.
For instance, for the \lstinline{globToRegExp} examples shown in
Listing~\ref{code:code-example-extracted-motivating-example} we obtain
the following observed candidate types
\begin{verbatim}
globToRegExp : (glob: string, opts: undefined) -> RegExp
globToRegExp : (glob: string, 
                opts: { extended : bool,
                        globstar?: any,
                        flags?: any }) -> RegExp
\end{verbatim}
After expanding manually the code base with one particular case, we obtain the following type
\begin{verbatim}
re = globToRegExp("*/www/{*.js,*.html}", {
  flags: "i",
  globstar: true,
});

globToRegExp : (glob: string, 
                opts: { extended? : any,
                        globstar: bool,
                        flags: string }) -> RegExp
\end{verbatim}

As the type for \lstinline{glob} and the return type
\lstinline{RegExp} are the same, we collapse the type for
\lstinline{opts} into a union type. The union with the
\lstinline{undefined} type is special and converted into an optional
argument so we arrive at a single collapsed candidate type
\begin{verbatim}
globToRegExp : (glob: string, 
                opts?: { extended? : bool,
                         globstar?: bool,
                         flags?: string }) -> RegExp
\end{verbatim}
We proceed in this way with merging further candidate types until we
reach a set of non-mergable types for the function. If this set
contains more than one candidate type, then we generate an overloaded
definition. For example, the candidate types
\begin{verbatim}
f(string): string
f(number): number
f(boolean): boolean
\end{verbatim}
cannot be merged as the return types are different. They give rise to
an overloaded definition with three alternatives. However, given the candidate types
\begin{verbatim}
g(string): string
g(number): string
g(boolean): boolean
\end{verbatim}
we can merge the first two types on the first argument (because all
remaining arguments and the return types are equal) by collecting \lstinline{string} and
\lstinline{number} into a union type. The resulting set of non-mergable types
is thus
\begin{verbatim}
g(string|number): string
g(boolean): boolean
\end{verbatim}

>>>>

\section{Motivating Example}
\label{sec:motivating-example}
The \NPM{} package \texttt{glob-to-regexp} provides functionality to turn a
glob expression for matching filenames in the shell into a regular
expression\footnote{\url{https://www.npmjs.com/package/glob-to-regexp}}. It 
has about 9.9M weekly downloads and 188 \NPM{} packages depend on it. If
a developer creates or extends TypeScript code that depends on the
\texttt{glob-to-regexp} library, the TypeScript
compiler and IDE require a declaration file for that library to
perform static checking and code completion, respectively. With
\texttt{dts-generate} we automatically generate a TypeScript
declaration file for \texttt{glob-to-regexp}. Our tool downloads the \NPM{} 
package, runs the examples extracted from its documentation, gathers
run-time information, and generates a TypeScript declaration
file. As the package is insufficiently documented, we provide one
additional example to generate the result 
shown in \figref{fig:motivating-example-glob-to-regexp-vscode} (see
discussion in Section~\ref{sec:typescr-decl-file}). This declaration is
ready for use in a TypeScript project:  Visual Studio's
code completion runs properly with the generated file. If the
\texttt{glob-to-regexp} package gets modified in the future, a new declaration
file can be generated automatically using
\texttt{dts-generate}. Our comparator tool (Section~\ref{sec:dts-compare}) can compare the new file
for incompatibilities with the previous declaration file.

@@

To construct the type signature for a function, we inspect the
primitive types of each argument as well as the interface created from
interactions on the argument in the function body. We do the same for
the return type of the function. This procedure yields a candidate
type for each function call observed during the execution.
For instance, for the \lstinline{globToRegExp} examples shown in
Listing~\ref{code:code-example-extracted-motivating-example} we obtain
the following observed candidate types
\begin{verbatim}
globToRegExp : (glob: string, opts: undefined) -> RegExp
globToRegExp : (glob: string, 
                opts: { extended: bool,
                        globstar: undefined,
                        flags: undefined }) -> RegExp
\end{verbatim}
These types are correct, but underapproximate the best possible type
for the package. Our comparison tool (see Section~\ref{sec:dts-compare}) classifies the differences as
solvable, hence we provide one more code example:
\begin{verbatim}
re = globToRegExp("*/www/{*.js,*.html}", {
  flags: "i",
  globstar: true,
});
\end{verbatim}
From this additional example, we obtain the following type:
\begin{verbatim}
globToRegExp : (glob: string, 
                opts: { extended: undefined,
                        globstar: bool,
                        flags: string }) -> RegExp
\end{verbatim}

As the type for the argument \lstinline{glob} and the return type
\lstinline{RegExp} are the same, we collapse the type for
\lstinline{opts} into a union type. The union with the
\lstinline{undefined} type is special and converted into an optional
argument. Moreover, the union of object types proceeds
componentwise. Altogether, we arrive at a single collapsed candidate
type.
\begin{verbatim}
globToRegExp : (glob: string, 
                opts?: { extended?: bool,
                         globstar?: bool,
                         flags?: string }) -> RegExp
\end{verbatim}
We proceed in this way with merging further candidate types until we
reach a set of non-mergable types for the function. If this set
contains more than one candidate type, then we generate an overloaded
definition. For example, the candidate types
\begin{verbatim}
f(string): string
f(number): number
f(boolean): boolean
\end{verbatim}
cannot be merged as the return types are different. They give rise to
an overloaded definition with three alternatives. However, given the candidate types
\begin{verbatim}
g(string): string
g(number): string
g(boolean): boolean
\end{verbatim}
we can merge the first two types on the first argument (because all
remaining arguments and the return types are equal) by collecting \lstinline{string} and
\lstinline{number} into a union type. The resulting set of non-mergable types
has two elements.
\begin{verbatim}
g(string|number): string
g(boolean): boolean
\end{verbatim}
